\section{Normalization}
Consider the proof of normalization (Theorem 12.1.6 in TAPL) for STLC with base types.
Extend this proof to STLC with any of the extension defined in the book. 
Do not use the unit type, as the only new term $\mathtt{unit}$ is already in normal form and thus the proof is trivial.\\
\emph{Hint: extend defninitoin 12.1.2 and lammas 12.1.3-12.1.5 using the extended language, then the normalization theorem follows}

\section{Bidirectional Typing}
Consider the system for Bidirectional Typing of STLC from the lecture, with the following rules for applications
\begin{minipage}{\textwidth}
  \begin{minipage}{0.45\textwidth}
    \begin{prooftree}
      \AxiomC{$\Gamma\vdash t_1 \Rightarrow T_{11}\rightarrow T_{12}$}
      \AxiomC{$\Gamma\vdash t_2\Leftarrow T_{11}$}
      \RightLabel{$\Rightarrow$\textsc{App}}
      \BinaryInfC{$\Gamma\vdash t_1\ t_2 \Rightarrow T_{12}$}
    \end{prooftree}
    \hfill
  \end{minipage}
  \begin{minipage}{0.45\textwidth}
    \begin{prooftree}
      \AxiomC{$\Gamma\vdash t_2 \Rightarrow T_{11}$}
      \AxiomC{$\Gamma\vdash t_1 \Leftarrow T_{11}\rightarrow T_{12}$}
      \RightLabel{$\Leftarrow$\textsc{App}}
      \BinaryInfC{$\Gamma\vdash t_1\ t_2 \Leftarrow T_{12}$}
    \end{prooftree}
    \hfill
  \end{minipage}
  \hfill
  \vspace{2em}
\end{minipage}
\begin{enumerate}
  \item Give a typing derivation for the following term, once for each variant for \textsc{App}. Add type ascriptions where necessary\\
    $$ (\lambda f. \lambda x. f\ x)\ (\lambda y.y)\ \mathtt{unit}$$
  \item Pick an extension from TAPL 11.5-11.12. For each new typing rule
    \begin{enumerate}
      \item State if the rule can be used for Inference, Checking or both in the bidirectional system (without adjusting the rule)
      \item Write down the corresponding bidirectional rule(s)
    \end{enumerate}
\end{enumerate}

\section{Constraint-based Typing}
\subsubsection{}

\begin{enumerate}
  \item Implement type reconstruction with Constraint-based Typing for STLC with the unit type 
    (without type annotations on lambdas) as shown in the lecture.
    Templates in Scala and rust will be provied in the forum.
  \item Choose one extension from TAPL chapter 11, and propose Constraint-Typing rules for it.
  \item Extend your type inference algorithm with the rules you came up with above
\end{enumerate}

\section{Reading}
Read through chapters 12 and 13 of TAPL and post your questions in the dedicated forum thread. 
You can also ask questions about bidirectional and constraint based inference dicussed only in the lecture.
